---
title: "RavenR Network Plots"
author: "Robert Chlumsky, Dr. James R. Craig"
date: "April 2020"
output: pdf_document
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{RavenR Mapping}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**This tutorial is currently under construction, and will be made fully available with a future version release**

# Introduction

This document is intended to discuss spatial mapping with RavenR, including animation of maps.

Some knowledge of R is presumed in this document. If you are not comfortable with R, take a look at any number of R training and Introductory resources, such as the [tRaining repository](https://github.com/rchlumsk/tRaining)) or the R training materials maintained by Dr. Kevin Shook under the [Centre for Hydrology](https://github.com/CentreForHydrology), both available on Github. 

Sections of this tutorial will use sample files from the Nith River basin, which is referenced in the Raven Tutorial files. The Nith river model can be downloaded from the  [Raven Tutorial #2](http://www.civil.uwaterloo.ca/jrcraig/Raven/Downloads.html).

## Load the RavenR library

Load the RavenR library from the console and view its contents with the following commands:

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(RavenR)
```

* plot network plot
* watershed plotting
* plot custom state variable output to basins
* animate the plots


Work in progress - workflow for network plots.

```{r}
# ProcessConnections.R

# test routine - reads resource files, stores as dataframes

# read in .rvi file process list (make separate routine - read in data frame here)
# read resource.dat file (most of the work). Store as data frame
# [process algorithm_name from_i to_i] - this should be native format of connection table RavenResource.dat
# RavenResource.dat table
# RavenAlgParams.dat table
# [processalgorithm parameter paramclass]
# RavenParameters.dat ->
# [param paramclass units autogen default min max {log?}]
# build list of connections - should just be appending sublists from resource.dat

#should specify directory of process connections
setwd("C:/James/Software/RavenR/testing/")
filename<-"RavenProcessConnections.dat"

HPTable<-rvi.read("Alouette.rvi")$HydProcTable
connections<-rvi.connections(HPTable,filename)
rvi.HPNetworkPlot(connections,"network-Alouette.pdf")
# this list can be sent to network plot as From/To connections


```


```{r}
# ProcessNetwork.R


#library
library(igraph)
#library(igraphinshiny)
source<-c("ATMOSPHERE","PONDED_WATER","PONDED_WATER","SOIL[0]","SOIL[0]","SOIL[1]")
target<-c("PONDED_WATER","SURFACE_WATER","SOIL[0]","SOIL[1]","SURFACE_WATER","SURFACE_WATER")
process<-c("Precipitation","Runoff","Infiltration","Percolation","Interflow","Baseflow")

# in real case -
# read in .rvi file process list (make separate routine - read in data frame here)
# read resource.dat file (most of the work). Store as data frame
# [process algorithm_name from_i to_i] - this should be native format of connection table RavenResource.dat
# RavenResource.dat table
# RavenAlgParams.dat table
# [processalgorithm parameter paramclass]
# RavenParameters.dat ->
# [param paramclass units autogen default min max {log?}]
# build list of connections - should just be appending sublists from resource.dat

nodes<-data.frame(source,target,process)
network<-graph_from_data_frame(d=nodes,directed=T)

plot.igraph(network,
            layout=layout_nicely(network),
            edge.color="black",
            edge.arrow.size=0.5,
            edge.label = nodes$process,
            edge.label.family="sans",
            edge.label.cex=0.7,
            #edge.curved=T,
            frame.color="darkslategray2",
            vertex.label.family="sans", #not working...
            vertex.shape="rectangle",
            vertex.size=120,
            vertex.size2=22,
            vertex.label.cex=0.7,
            vertex.color="darkslategray2")

```

