---
title: "RavenR Spatial Mapping Vignette"
author: "Robert Chlumsky, Dr. James R. Craig"
date: "March 2020"
output: pdf_document
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{RavenR Mapping}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**This tutorial is currently under construction, and will be made fully available with a future version release**

# Introduction

This document is intended to discuss spatial mapping with RavenR, including animation of maps.

Some knowledge of R is presumed in this document. If you are not comfortable with R, take a look at any number of R training and Introductory resources, such as the [tRaining repository](https://github.com/rchlumsk/tRaining)) or the R training materials maintained by Dr. Kevin Shook under the [Centre for Hydrology](https://github.com/CentreForHydrology), both available on Github. 

Sections of this tutorial will use sample files from the Nith River basin, which is referenced in the Raven Tutorial files. The Nith river model can be downloaded from the  [Raven Tutorial #2](http://www.civil.uwaterloo.ca/jrcraig/Raven/Downloads.html).

## Load the RavenR library

Load the RavenR library from the console and view its contents with the following commands:

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(RavenR)
```

* plot network plot
* watershed plotting
* plot custom state variable output to basins
* animate the plots


Work in the old SB.animateplot function

```{r}
SBMap.animate <- function(shpfilename,subIDcol,plot.daterange,cust.data,leg.title='Legend',leg.pos='bottomleft',
                               normalize.data=TRUE,colour1="azure",colour2="white",
                               num.classes=5,invalid.stop=TRUE,basins.label='subID',plot.title='',plot.invalid=F,
                               gif.filename='subbasin_animated_plot.gif',
                               gif.speed=1,
                               cleanup=T) {

  current.wd <- getwd()
  rand.dir <- paste0('scratch_SBMap',"_",paste0(ceiling(runif(20,min=0,max=10)),collapse=""))
  dir.create(rand.dir)

  # get the dates to plot from the supplied plot.daterange object
  plot.dates <- lubridate::date(cust.data[plot.daterange])

  for (i in seq(1,length(plot.dates))) {
    png(file=paste0(rand.dir,"/",sprintf("plot_%02d.png",i)), width=500, height=500)
    SBMap.plot(shpfilename,subIDcol,plot.dates[i],cust.data,leg.title,leg.pos,
               normalize.data,colour1,colour2,
               num.classes,invalid.stop,basins.label,plot.title, plot.invalid)
    dev.off()
  }
  try(dev.off(), silent=TRUE)

  # ---- OLD METHOD BELOW
  # convert the .png files to one .gif file using ImageMagick.
  # The system() function executes the command as if it was done
  # in the terminal. the -delay flag sets the time between showing
  # the frames, i.e. the speed of the animation.
  # system("convert -delay 80 *.png example_1.gif")
  # system(paste("convert -delay",as.character(transition.speed),paste0(rand.dir,"/","*.png"),gif.filename))
  # ------------


  # generate an animation from the images created

  pp <- list.files(pattern='*\\.png', recursive=TRUE)
  pp <- pp[grep(rand.dir,pp)]
  img <- image_read(pp)
  image_write_gif(img, path=gif.filename, delay=gif.speed)

  # delete subfolder
  if (cleanup) { unlink(rand.dir,recursive = T,force = T) }

  return(TRUE)
}
```


#' @param shpfilename filename of shapefile containing HRU polygons, with one column inidicating Raven HRU ID
#' @param subIDcol string of subbasin ID column in shapefile
#' @param plot.daterange string of date range to create in GIF
#' @param cust.data custom data set as read in by custom.read, for daily by_subbasin data
#' @param leg.title text for legend title
#' @param leg.pos position of legend
#' @param normalize.data whether to normalize data by all cust.data (TRUE) or just the data for the given date (FALSE)
#' @param colour1 string indicating which colour (text or code) to use on lower bound of range
#' @param colour2 string indicating which colour (text or code) to use on upper bound of range
#' @param num.classes number of classes to use in legend. Does not change the actual display colours
#' @param invalid.stop whether to stop if invalid basins are found (TRUE) or just continue with a warning (FALSE)
#' @param basins.label label to put on basins, one of c('None,'subID','value') to show nothing, subbasinIDs, or actual plotted values
#' @param plot.title title across top of plot
#' @param plot.invalid boolean indicating whether to plot invalid basins in grey (currently disabled)
#' @param gif.filename filename of outputted gif file
#' @param gif.speed time in seconds between images
#' @param cleanup boolean indicates whether to remove the scratch directory with image files stored.

